A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-18 paper.tex
Front-load ARG definitions

A genome ARG (gARG) is a graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
% This has two purposes: clarify that we're not talking about Shipilina
% et als edges, and also get "Graph theory" in here
(Note that we are using the term ``edge'' in its standard
graph-theoretic sense
and not the specialised interpretation of
\cite{shipilina2023origin}.)
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
% TODO we should change the notation for inheritance intervals here,
% as we use I to index individuals.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see section XXX for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
which genome $c$ inherits from $p$. Thus, each topological connection between
 a parent and child in the graph is ``annotated'' with set of
inheritance intervals $I$.

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local
% TODO do we mention "marginal" anywhere? Might not be worth the
% effort, surely "local" is clear?
is somewhat subtle (see section XXX), and it useful to have
a concrete definition and process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. The set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs model the inheritance between specific genomes,
annotating the edges of the graph with the intervals
of genetic inheritance. It is primarily concerned with
describing relationships and the details of inheritance
between pairs of genomes.
In contrast, the classical ARG data structure is defined
by evolutionary \emph{events},
and we therefore refer to it as an ``event ARG'' (eARG).
In an eARG we usually have two different types
of node,
representing common ancestor and recombination events that occurred in the
history of some set of sample genomes
and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use the standard convention here
that the order of parents
is significant, but additional information can be incorporated to
distinguish left and right parents
\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.

>>>>

A genome ARG (gARG) is a graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
% This has two purposes: clarify that we're not talking about Shipilina
% et als edges, and also get "Graph theory" in here
(Note that we are using the term ``edge'' in its standard
graph-theoretic sense
and not the specialised interpretation of
\cite{shipilina2023origin}.)
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
% TODO we should change the notation for inheritance intervals here,
% as we use I to index individuals.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see section XXX for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
%$[\ell, r)$
over
which genome $c$ inherits from $p$. Thus, each topological connection between
 a parent and child in the graph is ``annotated'' with set of
inheritance intervals $I$.

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
% TODO do we mention "marginal" anywhere? Might not be worth the
% effort, surely "local" is clear?
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs model the inheritance between specific genomes,
annotating the edges of the graph with the intervals
of genetic inheritance. It is primarily concerned with
describing relationships and the details of inheritance
between pairs of genomes.
In contrast, the classical ARG data structure is defined
by evolutionary \emph{events},
and we therefore refer to it as an ``event ARG'' (eARG).
The difference between gARGs and eARGs is subtle and the distinctions
at times rather porous. In essense, our main goal of introducing the
term ``Event ARG'' is to contrast the flexibility of the genome ARG
view with current practice. It is therefore useful to consider
the definition of an ARG from recent work.

% Todo this is rough. Ideally we want a bit of commentary running through
% the quotations.
\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
\citet{shipilina2023origin} emphasise the reality  ``The ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events''. In contrast,
\citet{brandt2021evaluation} tie the definition of an ARG explicitly
to a generative process:
``The full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]

The common features of these definitions is that they are expressed
in terms of evolutionary events, emphasise the exhaustiveness
of the record (full, complete), and they do not specify the mechanism
by which partial genetic inheritance occurs. Just as the graph topology
in a gARG defines the paths along which genetic inheritance occurs, some
additional information is required to define the actual regions of genome
inherited.

% % The full ancestral recombination graph (ARG) is a structure that encodes all
% % coalescence and recombination events resulting from the stochastic process of
% % the coalescent with recombination.

% historial trend discussed in the XXX section).
% \citet{rasmussen2022espalier} have a less restrictive definition, in
% which a ``full'' ARG is one that contains complete information about
% recombination events, without necessarily being tied to a particular
% stochastic process.
% While [something positive], it is important to be clear on
% what the overall goal of such precise inferences are,
% what the limitations on when they can be made are,
% and their downstream utility in applications.

In an eARG we usually have two different types
of node,
representing common ancestor and recombination events that occurred in the
history of some set of sample genomes
and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use the standard convention here
that the order of parents
is significant, but additional information can be incorporated to
distinguish left and right parents
\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.
